\begin{center}
{\bf {\Large 
		Credits
            %=============%
}}
\end{center}

% Apologies to anyone I left out!  It wasn't intentional ... TLS.

\begin{quote}
Day-to-day care and feeding of XSB including bug fixes and ports, was
led by Kostis Sagonas, and aided by David Warren, Terrance Swift,
Prasad Rao, Steve Dawson, Juliana Freire, and Baoqiu Cui.

To date (\version), the core engine development of the SLG-WAM has
been mainly implemented by Terrance Swift, Kostis Sagonas, Prasad Rao
and Juliana Freire.  The breakdown, roughly, was that Terrance Swift
wrote the initial tabling engine, Prasad Rao wrote the trie-based
table access routines, and Kostis Sagonas implemented most of tabled
negation.  Juliana Freire revised the table scheduling mechanism
starting from Version 1.5.0 to create a more efficient engine, and
implemented the engine for local evaluation.  David Warren and Prasad
Rao implemented the aggregate code.  The {\tt auto\_table} and {\tt
suppl\_table} directives were written by Kostis Sagonas.  Memory
management code and the command line execution interface were written
by Ernie Johnson, Prasad Rao and Bart Demoen.  Rui Marques improved
the trailing of the SLG-WAM and implemented the port for 64-bit
architectures.  The safety check for tabling within the scope of cuts
was written by Kate Dvortsova.

Of other features in the system, Kostis Sagonas added HiLog
compilation to XSB, along with the associated builtins.  Steve Dawson
implemented Unification Factoring.  The DCG expansion module was
written by Kostis Sagonas.  The multifile directive was written by
Baoqiu Cui.  Assert and retract code was written by David Warren,
significantly revised by Rui Marques, and based originally on code
written by Jiyang Xu.  Trie assert code was written by Prasad Rao.
The interface from C to Prolog was implemented by David Warren as was
the DLL interface; the interface from Prolog to C (foreign language
interface) was developed by Jiyang Xu, Kostis Sagonas and Steve
Dawson.  The Oracle interface was written by Hassan Davulcu and Ernie
Johnson.  The ODBC interface was written by Lily Dong.
The current version of {\tt findall/3} was written by Bart Demoen.
C.R. Ramakrishnan wrote the mode analyzer for XSB.  The WFSX libraries
were written by Luis Pereira and Jose Alferes.

Harald Schroepfer helped the XSB group with the Solaris port, and
Yiorgos Adamopoulos suggested the bits to use for the HP-700 series
port.  Steven Dawson, Larry B. Daniel and Franklin Chen were
responsible for the MkLinux and Solaris x86 ports.

Thanks are also due to Weidong Chen for his early work on Prolog
clause indexing and Richard O'Keefe, who contributed the Prolog
code for the Prolog reader and the C code for the tokenizer.  

The starting point of \ourprolog\ was PSB-Prolog 2.0 by Jiyang Xu.
PSB-Prolog in its turn was based on SB-Prolog, primarily designed and
written by Saumya Debray, David S. Warren, and Jiyang Xu.  This manual
was mostly written by Kostis Sagonas and Terrance Swift except for the
chapter on the Oracle interface written by Hassan Davulcu.

\end{quote}

