

\newenvironment{qrules}{\begin{quote}\sf\begin{tabular}[t]{l}}%
{\end{tabular}\end{quote}}


\newcommand{\obj}{\textit{obj}}
\newcommand{\db}[1]{\ensuremath{\mathcal{#1}}}

\newcommand{\xany}{\textsf{any}}

\newcommand{\xplus}{\ensuremath{^+}}
\newcommand{\xstar}{\ensuremath{^*}}
\newcommand{\xinv}{\ensuremath{^{-1}}}
\newcommand{\xopt}{\ensuremath{^{?}}}

\newcommand{\xto}[1]{\ensuremath{^{#1}}}
\newcommand{\xcond}[1]{\ensuremath{\textsf{if}(#1)}}
\newcommand{\xif}[1]{\ensuremath{\textsf{if}(#1)}}
\newcommand{\xmu}[1]{\ensuremath{\tcmu(#1)}}
\newcommand{\xmuif}[2]{\ensuremath{\tcmu(#1,#2)}}


\newcommand{\xconc}{\ensuremath{{\cdot}}}
\newcommand{\xor}{\ensuremath{|}}

\newcommand{\nnot}{\mbox{$\neg$}}                           % negation
\newcommand{\query}{\mbox{$\, ?\! - \, $}}                  % query
\newcommand{\impl}                                          % implication
  {\mbox{\Large $\; {\bf \leftarrow} \;$}}  
\newcommand{\isa}{\,{\bf{:}}\,}
\newcommand{\subcl}{\,{\bf{::}}\,}
\newcommand{\eq}{\ensuremath{\doteq}}                           % equation

% f-logic arrows

\newcommand{\fd}{\ensuremath{{\rightarrow}}}                   % scalar
\newcommand{\bfd}{\ensuremath{{\bullet\!\!\!\fd}}}            % " + inheritable
\newcommand{\mvd}{\ensuremath{{\rightarrow\!\!\!\!\rightarrow}}}  % multivalued
\newcommand{\bmvd}{\ensuremath{{\bullet\!\!\!\mvd}}}              % " + inheritable
\newcommand{\Fd}{\ensuremath{{\Rightarrow}}}                      % scalar signature
\newcommand{\Mvd}{\ensuremath{{\Rightarrow\!\!\!\!\Rightarrow}}}  % multiv signature



% curved f-logic arrows

\newcommand{\anyd}{\ensuremath{\leadsto}}                       % non-inheritable
\newcommand{\bleadsto}{\ensuremath{\bullet\!\!\!\leadsto}}     % inheritable
\newcommand{\banyd}{\bleadsto}                              % "
\newcommand{\Leadsto}{\ensuremath{\approx}\!\!{>}}            % signature
\newcommand{\Anyd}{\Leadsto}                                % "

\newcommand{\FdConstr}{\ensuremath{\stackrel{constr}{\Fd}}}
\newcommand{\MvdConstr}{\ensuremath{\stackrel{constr}{\Mvd}}}

\newlength{\flogicindent}


\newlength{\flength}
\newlength{\counterlength}


\newcommand{\la}{\ensuremath{\,\leftarrow\,}}

\newcommand{\anon}{\_}

\newcommand{\note}[1]{\textit{[[#1]]}}
\newcommand{\nterm}[1]{\ensuremath{\langle}\textit{#1}\ensuremath{\rangle}}


\newcommand{\NI}{\noindent}

\newcommand{\bs}{\ensuremath{\backslash}}
\newcommand{\FLIP}{{\mbox{\sc Flip}}\xspace}
\newcommand{\FLORID}{{\textsc{Florid}}\xspace}
\newcommand{\fl}{{F-logic}\xspace}


\newcommand{\consts}{\ensuremath{\mathcal{C}}}
\newcommand{\funcs}{\ensuremath{\mathcal{F}}}
\newcommand{\preds}{\ensuremath{\mathcal{P}}}
\newcommand{\vars}{\ensuremath{\mathcal{V}}}

\newcommand{\HU}{\ensuremath{U}}
\newcommand{\HB}{\ensuremath{\mathcal{HB}}}
\newcommand{\ext}{\ensuremath{^{\star}}}




\chapter{Tour de \FLIP}

\begin{center}
{\Large {\bf By Bertram Lud\"ascher}}
\end{center}

\section{Introduction}

\FLIP is a simple F-logic to XSB compiler. It takes a program written in the 
F-logic language \cite{KLW95} (which must be in a file with extension {\tt
  .flp}, {\it e.g.}, {\tt file.flp}) and outputs a file with extension {\tt
  .P} ({\it e.g.}, {\tt file.P}), which is a regular XSB program. This
program is then passed to XSB for compilation (yielding {\tt file.O}) and
execution. 

\FLIP is part of the official distribution of XSB beginning with version
2.0. It is organized as an XSB package and lives in the directory
%%
\begin{quote}
 \verb|<xsb-installation-directory>/packages/flip/|  
\end{quote}
%%
It comes with a number of demo programs that live in
%%
\begin{quote}
 \verb|<xsb-installation-directory>/packages/flip/demos/|  
\end{quote}
%%
The demos can be run by issuing the command
\verb|rundemo("demo-filename").| at the \FLIP prompt, {\it e.g.},
\verb|rundemo("flogic_basics").|

To start using \FLIP, you must first invoke XSB and then load the {\tt flip} 
module:
%%
\begin{quote}
  \tt
foo>~~xsb  \\
\tt
[ XSB chat during loading deleted ]\\
\tt
| ?- [flip].\\
\tt
[flip loaded]\\
\tt
[flpshell loaded]\\
\tt
| ?-
\end{quote}
%%
At this point, it is possible to use some of the \FLIP shell commands, such as
running demos. (The only trick is that you must append ``\_'' to every \FLIP
shell command.) However, you are still not inside \FLIP command loop, so you 
cannot run F-logic queries interactively. In order to do that, \FLIP shell
must be started:
%%
\begin{quote}
  \tt
| ?- flip\_shell.  \\
 \tt
[ FLIP chat deleted ] \\
 \tt
FLIP> ?-
\end{quote}
%%

At this stage, \FLIP interpreter takes over and F-logic syntax becomes the
norm. To get back to the XSB command loop, type {\tt Control-D} or 
%%
\begin{quote}
  \tt
| ?- end.  
\end{quote}
%%


\section{\FLIP Shell Commands}

\FLIP commands:
\begin{verbatim}
  help          :       show this info
  parse(FILE)   :       parse "FILE.flp"; create "FILE.P"
  run(FILE)     :       compile "FILE.flp"; execute "FILE.P"
  rundemo(FILE) :       same as run("FILE"), but "FILE.flp"
                        is taken from the FLIP demo directory
  load(FILE)    :       load and execute "FILE.O"
  all           :       show all solutions (default)
  one           :       show solutions one after another
  end           :       say CIAO to FLIP
  call(GOAL)    :       call underlying Prolog
\end{verbatim}

The commands {\tt parse(FILE)}, {\tt run(FILE)}, and {\tt load(FILE)} all
use the XSB {\tt library\_directory} predicate to search for the module
that is passed to them as an argument. The command {\tt rundemo(FILE)} is
similar, except that it first looks for {\tt FILE} in the \FLIP demo
directory. 


\section{F-logic and \FLIP by Example}

At some point in the future, this section will contain a number of small
introductory examples of the use of F-logic and \FLIP. Meanwhile, the
reader is referred to the excellent tutorial written by the members of the
\FLORID project, which is yet another implementation of \fl.\footnote{
  %%
  See {\tt http://www.informatik.uni-freiburg.de/$\sim$dbis/florid/} for more
  details.
  %%
  }
%%
Since \FLIP
and \FLORID share the same syntax, the examples in that tutorial are
also valid programs in \FLIP.



\section{Inside \FLIP}


\subsection{Modules}

\FLIP consists of the following modules:
\begin{itemize}
\item \texttt{flpshell.P}: Top-level module which provides the \FLIP shell
  commands, i.e., for loading and executing \FLIP programs
  (\texttt{load/1}, \texttt{run/1}), setting the output mode
  (\texttt{all/0} or \texttt{one/0} solution(s) at a time) and -- last
  not least -- for directly issuing queries against the loaded
  database/program.
\item \texttt{flptokens.P}: \FLIP tokenizer, based on R. O'Keefe's public
  domain tokenizer.
\item \texttt{flpparser.P}: DCG parser for F-logic with path expressions
  in the body.
\item \texttt{flpcompiler.P}: Implements the compilation of F-logic to
  XSB.
\item \texttt{flputils.P}: Miscellaneous utility predicates.
\end{itemize}



\subsection{How \FLIP Works}



\paragraph{Overview.}

As an F-logic to XSB compiler, \FLIP first parses its argument file and then
compiles it to XSB syntax. For instance the command
\begin{verbatim}
        FLIP> ?- run("myprog").  
\end{verbatim}
first parses the program \verb|"myprog.flp"| (using \texttt{flpparser.P})
and then creates the XSB file \verb|"myprog.P"| (using
\texttt{flpcompiler.P}).  Look at this file to see what has become of your
F-logic program! The compilation consists mainly of a flattening procedure
sketched below.  Next, \verb|"myprog.P"| is compiled by XSB, yielding
\verb|"myprog.O"| which is then executed.  If \verb|"myprog.flp"| contains
queries, these are immediately executed by XSB (provided there
are no errors).

The main purpose of the \FLIP shell, however, is to allow the
evaluation of ad-hoc F-logic queries. For example, after having
requested the execution of the \texttt{default.flp} file from the demo
directory (using the
command \texttt{FLIP>~?-rundemo("default").}), you may ask
\begin{verbatim}
    FLIP> ?-  X..kids[                  % Whose kids
                 self -> K;             % ... (list them by name)
                 hobbies ->>            % ... have hobbies
                 {H:dangerous_hobby}    % ... that are dangerous?
    ]. 
\end{verbatim}
\FLIP will parse, flatten, and evaluate this query in the same way as
the queries in a source file.





\paragraph{Flattening F-logic.}

The core of the \FLIP system is the predicate
\texttt{comp(+Flogic\_rule,-Prolog\_rules)}, which rewrites a complex
F-logic rule into a set of ``flattened'' rules, which in turn can be
directly mapped to XSB.

Consider, e.g., the following complex F-logic molecule, representing
facts about the object \texttt{mary} (the syntax of F-logic is given in
Section \ref{sec-basic-flogic}):

\begin{quote}
{\small\begin{verbatim}
mary:employee[age->29;kids->>{tim,leo};salary@(1998)->a_lot].
\end{verbatim}}
\end{quote}

Clearly, such a complex molecule can be easily flattened into a conjunction
of F-logic atoms.  For the different kinds of F-logic atoms we use a set of
corresponding Prolog atoms. E.g., the result of translating the above
F-molecule will be:

\begin{quote}
{\small \begin{verbatim}
isa_(mary,employee).                    % mary:employee.
fd_(mary,@(age),29).                    % mary[age->29].
mvd_(mary,@(kids),tim).                 % mary[kids->>{tim}].
mvd_(mary,@(kids),leo).                 % mary[kids->>{leo}].
fd_(mary,@(salary,1998),a_lot).         % mary[salary@(1998)->a_lot].
\end{verbatim}}
\end{quote}

%% For further details on path expressions see Section \ref{sec-references}.


\paragraph{Closure Axioms.}

In addition to the facts and rules given by the user, there are additional
rules, called \emph{closure axioms}, which have to be considered when
evaluating an F-logic program. In the current version, \FLIP implements
the closure axioms by appending the file \texttt{lib/flptrailer.P} to every
{\tt .P} file it generates. In this way, the closure rules get appended to
every user program. These rules capture:

\begin{itemize}
\item The closure of ``\subcl'', i.e., the subclass hierarchy.  A
  runtime check warns about cycles in the subclass hierarchy.
\item The closure of ``isa'', i.e., if $X\isa C, C\subcl D$ then
  $X\isa D$. 
\item Monotone inheritance of ``\bfd'' and ``\bmvd'' methods,
  e.g., if $C[M@(X)\bfd R],O\isa C$ then $O[M@(X)\fd R]$.
\end{itemize}

Additionally, \texttt{flptrailer.P} contains rules to check the
consistency of functional (single-valued) methods.

\paragraph{Known Bugs.}

\begin{itemize}
\item Rules like \texttt{p :- obj[].}  do not generate code.
\item Non-ground answers are not properly displayed, e.g.,
  \texttt{?-X::Y}.
\end{itemize}

\subsection{\FLIP vs FLORID}

The syntax of \FLIP and some of its design decisions are borrowed from
Florid, an F-logic interpreter developed at Freiburg University in Germany.
For more information on Florid please visit the project home page:
\verb|http://www.informatik.uni-freiburg.de/~dbis/florid/|. The following
is a list of differences with the Florid system.

\begin{itemize}
\item \FLORID
  \begin{itemize}
  \item (Semi-)naive bottom-up evaluation.
  \item Single-valued and multi-valued  path expressions in the body.
  \item Single-valued path expressions in the head for anonymous
    object creation.
  \item Nonmonotonic inheritance (trigger semantics).
  \item Derived equalities determined by functional methods.
  \item ``Hard-wired'' closure axioms.
  \item Additional Web-access features (access to Web pages via
    \texttt{get}-method, Perl regular expressions, Perl interface).
  \item Large C++ system.
  \end{itemize}
\item \FLIP
  \begin{itemize}
  \item Translation of F-logic into XSB rules.
  \item Top-down evaluation of the generated rules. If tables, the compiled 
    programs can be much more efficient than the corresponding Florid
    programs.
  \item Single-valued and multi-valued  path expressions in the body.
  \item No path expressions in the head. Currently, the grammar allows such 
    expressions, but the result is unpredictable.
  \item Monotonic inheritance.
  \item No derived equalities for functional methods---consistency check is 
    done instead.
  \item Closure axioms as additional rules.
  \item Small Prolog system.
  \item Full access to the underlying XSB system, including its database
    connectivity.
  \end{itemize}
\end{itemize}





\section{Syntax of \FLIP Programs (DCG)}

The DCG grammar below specifies the syntax of \FLIP programs.  Symbols
in brackets \texttt{[...]} correspond to {terminal symbols} as
returned by the tokenizer (\texttt{tokens.P}). For example, the rule
\begin{verbatim}
id_term                 -->     [identifier(_),'('], paths, [')'].
\end{verbatim}
stands for the (E)BNF rule:
\begin{quote}
  \nterm{id\_term} ::= \textsf{identifier} "\textsf{(}" \nterm{paths} "\textsf{)}"
\end{quote}
where \nterm{id\_term} and \nterm{paths} are nonterminals (defined by
the DCG grammar) and \textsf{identifier}, \textsf{(}, and \textsf{)}
are terminals (defined by the tokeniser). In the actual \FLIP parser
(\texttt{parser.P}), additional arguments to the DCG predicates are
used to propagate variable bindings from the rule body to the head.
For example, the actual code of the  above rule is
\begin{verbatim}
id_term(fn(F,L))        -->     [identifier(F),'('], paths(L), [')'].
\end{verbatim}
Hence, the actual value \texttt{F} of the identifier token (returned
by the tokeniser) and of the path list \texttt{L} (returned by the
DCG) are used to construct the term \texttt{fn(F,L)}
(denoting a function symbol \texttt{F} and a list of arguments
\texttt{L}). 

For better readability, arguments to nonterminals have been omitted
(for nonterminals) or anonymized (for terminals). A list of \emph{foo}
expressions is denoted by \emph{foo\textbf{s}}; an optional (i.e.,
possibly empty) \emph{foo} expression is named \emph{foo\textbf{1}}.
The DGC grappar of \FLIP is shown in Figure~\ref{fig-flip-dcg}.


\begin{figure}[htbp]
\begin{minipage}[t]{.48\textwidth}
{\scriptsize
\begin{verbatim}
rule                -->     head.
rule                -->     head, [atom(' :-')], body.
rule                -->     [atom(' ?-')], body.

head                -->     terms.

terms               -->     term, terms1.

terms1              -->     [','], terms.    
terms1              -->     [].

body                -->     literals.

literals            -->     literal, literals1.

literals1           -->     [','], literals.
literals1           -->     [].

literal             -->     term.
literal             -->     [identifier(not)], term.
literal             -->     [atom(' ~')], term.

term                -->     f_mol.
term                -->     p_term.              
                        
p_term              -->     [identifier(_)], args.
p_term              -->     path, special_pred, path.
                                
args                -->     ['('], paths, [')'].
args                -->     [].

special_pred        -->     [atom(' >')].
special_pred        -->     [atom(' <')].
special_pred        -->     [atom(' =')].
special_pred        -->     [atom(' >=')].
special_pred        -->     [atom(' =<')].
special_pred        -->     [atom(' =/=')].
                                                        
path                -->     reference, opt_specification.

f_mol               -->     reference, specification.

reference           -->     object, sms.
                          
sms                 -->     specification0, meth_ref, sms.
sms                 -->     []. 

object              -->     id_term.            
object              -->     ['('], path, [')'].

specification0      -->     specification.
specification0      -->     [].

specification       -->     meth_spec.
specification       -->     isa_spec, meth_spec1.

opt_specification   -->     specification.
opt_specification   -->     [].
\end{verbatim}}
\end{minipage}
\begin{minipage}[t]{.48\textwidth}
{\scriptsize
\begin{verbatim}
isa_spec            -->     isa_symbol, object.

isa_symbol          -->     [atom(' :')].
isa_symbol          -->     [atom(' ::')].

meth_spec           -->     ['['], meths, [']'].
meth_spec           -->     ['['], [']'].

meth_spec1          -->     meth_spec.
meth_spec1          -->     [].

meth_ref            -->     colon, meth_appl.

colon               -->     [atom(' ..')].
colon               -->     [atom(' .')].
colon               -->     [atom(' !!')].
colon               -->     [atom(' !')].

meth_appl           -->     object, meth_args.

meth_args           -->     [atom(' @'),'('], paths, [')'].
meth_args           -->     [].

meths               -->     meth_appl, meth_result, rest_meths.

rest_meths          -->     [atom(' ;')], meths.
rest_meths          -->     [].

paths               -->     path, paths1.
                                
paths1              -->     [','], paths.    
paths1              -->     [].

meth_result         -->     fun_arrow, path.

meth_result         -->     set_arrow, path.
meth_result         -->     set_arrow, ['{'], ['}'].
meth_result         -->     set_arrow, ['{'], paths, ['}'].

meth_result         -->     sig_arrow, path.
meth_result         -->     sig_arrow, ['('], [')'].
meth_result         -->     sig_arrow, ['('], paths, [')'].

fun_arrow           -->     [atom(' ->')].   
fun_arrow           -->     [atom(' *->')].   
set_arrow           -->     [atom(' ->>')].    
set_arrow           -->     [atom(' *->>')].
sig_arrow           -->     [atom(' =>')].        
sig_arrow           -->     [atom(' =>>')].    

id_term             -->     [identifier(_),'('], paths, [')'].
id_term             -->     [identifier(_)].
id_term             -->     [var(_,_)].
id_term             -->     [string(_)].
id_term             -->     [num(_)].
id_term             -->     [atom(' -'),num(_)].
\end{verbatim}}
\end{minipage}
\caption{\FLIP DCG (Definite Clause Grammar)}
\label{fig-flip-dcg}
\end{figure}



\section{Syntax of \fl\ and Path Expressions }

The following is adopted from \cite{ludaescher-himmeroeder-IS-98}.


\subsection{Basic F-logic Syntax}\label{sec-basic-flogic}


\begin{itemize}
\item \emph{Symbols}: The \fl\ alphabet fo \emph{object constructors}
  consists of the sets \funcs (function symbols), \preds (predicate symbols
  including $\eq$), and \vars (variables).  Variables are denoted by
  capitalized symbols (e.g., $X,\textit{Name}$), whereas all other symbols,
  especially constants (which are 0-ary object constructors) are denoted in
  lowercase (e.g., $a,\textit{john}$).  An expression is called
  \emph{ground} if it involves no variables.  In addition to the usual
  first-order connectives and symbols, there are a number of special
  symbols: ], [, \}, \{, \fd, \mvd, \Fd, \Mvd,
  \isa, \subcl.\footnote{
    %%
    We do not deal with inheritance in this manual, so we
    omit the symbols for \emph{inheritable} methods
    \cite{KLW95}.
    %%
    }
  %%
\item \emph{Id-Terms/Oids}:  \medskip
  
  \hfill (0)
  \begin{minipage}[t]{.80\textwidth}
    First-order terms over \funcs\ and \vars\ are called
    \emph{id-terms}, and are used to name objects, methods, and
    classes.  Ground id-terms correspond to \emph{logical object
      identifiers} (\emph{oid}s), also called object \emph{names}.
  \end{minipage}
  \hfill ~
\item \emph{Atoms}: Let $O,M,R_{i},X_{i},C,D,T$ be id-terms.  In
  addition to the usual first-order atoms, like $p(X_1,\dots,X_n)$, there
  are the following basic types of atoms: \medskip

  \begin{math}
    \hfill (1)~O[M\fd R_0] \hfill (2)~O[M\mvd \{R_1,\dots,R_n\}]
    \hfill (3)~C[M\Fd T] \hfill (4)~C[M\Mvd T]. \hfill
  \end{math} \medskip
  
  (1) and (2) are \emph{data atoms}, specifying that a \emph{method} $M$
  applied to an object $O$ yields the result object $R_i$. In (1), $M$ is a
  \emph{single-valued} (or \emph{scalar}) method, i.e., there is
  at most one $R_0$ such that $O[M\fd R_0]$ holds. In contrast, in
  (2), $M$ is \emph{multi-valued}, so there may be several result
  objects $R_i$. For $n=1$ the braces may be omitted.\\
                                %
  (3) and (4) denote \emph{signature atoms}, specifying that the
  (single-valued and multi-valued, respectively) method $M$ applied to
  objects of \emph{class} $C$ yields results of type $T$.
  
  The organization of objects in classes is specified by
  \emph{isa-atoms}: \medskip

  \begin{math}
    \hfill (5)~O\isa C \hfill (6)~C\subcl D. \hfill
  \end{math} \medskip

  (5) defines that $O$ is an \emph{instance} of class $C$, while (6)
  specifies that $C$ is a \emph{subclass} of $D$. 
\item \emph{Parameters}: Methods may be \emph{parameterized}, so
  \begin{math}
    M@(P_1,\dots,P_k)
  \end{math} is allowed in (1)--(4), where $P_1,\dots,P_k$ are
  id-terms; e.g., \textsf{john[salary@(1998)\fd 50000]}.
  
\item \emph{Programs}: \fl\ \emph{literals}, \emph{rules}, and
  \emph{programs} are defined as usual, based on \fl\ atoms.
\end{itemize}

\NI As a concise notation for several atoms specifying properties of the
same object, \emph{F-molecules} can be used. For instance, instead of
$\textsf{john:person}\land\textsf{john[age\fd
  31]}\land\textsf{john[children\mvd\{bob,mary\}]}$, we can simply write
\textsf{john\isa person[age\fd 31; children\mvd\{bob,mary\}]}.


\paragraph{Object Model.} 
An \emph{\fl\ database (instance)} is represented by a set of facts
(i.e., atoms and molecules). Object-oriented databases often admit a
natural graph-like representation, which is, e.g., exploited in GOOD
\cite{gyssens-paredaens-vdbussche-vgucht-TKDE-94}. Similarly, \fl\ 
databases can be represented as labeled graphs where nodes correspond
to logical oids, and where the different kinds of labeled edges (\fd,
\mvd, \Fd, \Mvd) are used to represent different relations among
objects (at the class or instance level).

\begin{example}
  {\bf (Publications Database)}
  Figure~\ref{fig-flogic-model} depicts an \fl representation of a fragment
  of an object-oriented publications database.
\end{example}


\begin{figure}[htbp]
\begin{tabular}{c}
\fbox{\small\sf
  ~\begin{tabular}{l}
    paper[authors\Mvd  person; title\Fd string].\qquad
    conf\_p\subcl paper. \qquad  journal\_p\subcl paper.\\
    journal\_p[in\_vol\Fd volume]. \qquad
    conf\_p[at\_conf\Fd conf\_proc].\\
    journal\_vol[of \Fd journal; volume\Fd integer; 
               number\Fd integer; year\Fd integer].\\  
    journal[name\Fd string; publisher\Fd string;
            editors@(integer)\Mvd person]. \\
    conf\_proc[of\_conf\Fd conf\_series; year\Fd integer;
               editors@(integer)\Mvd person]. \\
    conf\_series[name\Fd string]. \qquad publisher[name\Fd string].\\
    person[name\Fd string; affil@(integer)\Fd institution]. \qquad
    institution[name\Fd string; address\Fd string].\smallskip\\
    \hline
    $o_{j1}$\isa journal\_p[%
      title\fd ``Records, Relations, Sets, Entities, and Things'';
      authors\mvd$\{o_{mes}\}$; in\_vol\fd $o_{i11}$]. \\
    $o_{di}$\isa conf\_p[%
      title\fd ``DIAM II and Levels of Abstraction'';
      authors\mvd$\{o_{mes},o_{eba}\}$; at\_conf\fd $o_{v76}$]. \\
    $o_{i11}$\isa journal\_vol[of\fd $o_{is}$; number\fd 1; volume\fd 1; year\fd1975]. \\
    $o_{is}$\isa journal[name\fd``Information Systems''; editors@(...)\mvd $\{o_{mj}\}$]. \\
    $o_{v76}$\isa conf\_proc[of\fd vldb; year\fd 1976; editors\mvd $\{o_{pcl},o_{ejn}\}$].\\
    $o_{vldb}$\isa conf\_series[name\fd``Very Large Databases'']. \\
    $o_{mes}$\isa person[name\fd``Michael E. Senko'']. \quad
    $o_{mj}$\isa person[name\fd``Matthias Jarke''; affil@($\dots$)\fd $o_{rwt}$]. \\
    $o_{rwt}$\isa institution[name\fd``RWTH\_Aachen'']. \qquad \ldots
\end{tabular}~}
\end{tabular}
\caption{A Publications Object Base and its Schema Represented 
  Using \fl}\label{fig-flogic-model}
\end{figure}





\subsection{Path Expressions}

In addition to the basic \fl\ syntax, the \FLIP  system also supports
\emph{path expressions} to simplify object navigation along
single-valued and multi-valued method applications, and to avoid
explicit join conditions \cite{frohn-lausen-uphoff-VLDB-94}.  The
basic idea is to allow the following \emph{path expressions} wherever
id-terms are allowed:\footnote{
  %%
  \FLIP doesn't support path expressions in rule
  heads.
  %%
  }
%%

  \medskip

\begin{math}
  \hfill (7)~O.M \hfill (8)~O..M \hfill
\end{math} \medskip

\NI The path expression in (7) is \emph{single-valued}; it refers to
the unique object $R_0$ for which $O[M\fd R_0]$ holds; (8) is a
\emph{multi-valued} path expression; it refers to each $R_i$ for which
$O[M\mvd\{R_i\}]$ holds. $O$ and $M$ may be id-terms or path
expressions; moreover, $M$ may be parameterized, i.e., of the form
$M@(P_1,\dots,P_k)$.
  
In order to obtain a unique syntax and to specify different orders of
method applications, parentheses are used: By default, path
expressions associate to the left, so $a.b.c$ is equivalent to
$(a.b).c$ and specifies the unique object $o$ such that $a[b\fd x]
\land x[c\fd o]$ holds (note that $x=a.b$). In contrast, $a.(b.c)$ is
the object $o'$ such that $b[c\fd x'] \land a[x'\fd o']$ holds (here,
$x'=b.c$); generally, $o'\neq o$. Note that in $(a.b).c$, $b$ is a
method name, whereas in $a.(b.c)$ it is used as an object name.
Observe that function symbols can also be applied to path
expressions, since path expressions (like id-terms) are used to
reference objects.
  
As path expressions and F-logic atoms can be arbitrarily nested, this leads
to a concise and very flexible specification language for object
properties, as illustrated in the following example.

\begin{example}[Path Expressions]\label{Ex:PathExpr}
  Consider again the schema given in Figure~\ref{fig-flogic-model}.
  Given the name $n$ of a person, the following path expression
  references all editors of conferences in which $n$ had a
  paper:\footnote{Each occurrence of ``\_'' denotes a distinct
    don't-care variable (which is existentially quantified at the
    innermost level).}
\begin{qrules}
  \anon\isa conf\_p[authors\mvd\{\anon [name\fd $n$]\}].at\_conf..editors
\end{qrules}
Therefore, the answer to the \emph{query}
\begin{qrules}
  ?- P\isa conf\_p[authors\mvd\{\anon [name\fd
  $n$]\}].at\_conf[editors\mvd\{E\}].
\end{qrules}
is the set of all pairs (\textsf{P},\textsf{E}) such that \textsf{P}
is (the logical oid of) a paper written by $n$, and \textsf{E} is the
corresponding proceedings editor.  If one is also interested in the
affiliations of the above editors when the papers were published, we only
need to slightly modify our query:
\begin{qrules}
  ?- P\isa conf\_p[authors\mvd\{\anon [name\fd
  $n$]\}].at\_conf[year\fd Y]..editors[affil@(Y)\fd A].
\end{qrules}
\end{example}
Thus, \FLIP's path expressions support navigation and specification
of object properties along two dimensions: the ``depth'' dimension
corresponds to navigation along method applications (``.''  and
``..''), while the bracketed specification lists specify
properties of the intermediate objects (the ``breadth'' dimension). Note
that constraints \emph{within} the expressions can be stated using
variables.

To access intermediate objects that arise implicitly in the middle
of a path expression, one can define the method \textsf{self} a
$X[\textsf{self}\fd X]$ and then simply
write $\dots[\textsf{self}\fd O]\dots$ anywhere in a complex
path expression. This would bind the id of the current object to the
variable $O$.\footnote{
  %%
  A similar feature is used in other
  languages, e.g., XSQL \cite{xsql-92}.
  %%
  }
%%

\begin{example}[Path Expressions with \textsf{self}]\label{ex-path-self}
  Recall the second query in Example~\ref{Ex:PathExpr}. If the user is
  also interested in the respective conferences, the query can be
  reformulated as
\begin{qrules}
   ?- P\isa conf\_p[authors\mvd\{\anon [name\fd
   $n$]\}].at\_conf[self\fd C; year\fd Y]..editors[affil@(Y)\fd A]. 
\end{qrules}
\end{example}

\subsection{References: Truth Value vs.\ Object Value}\label{sec-references}

Id-terms, F-logic atoms, and path expressions can all be used to
reference objects. This is obvious for id-terms (0) and path
expressions (7--8). Similarly, F-logic atoms (1--6) have not only a
truth value, but also reference objects, i.e., yield an object value.
For example, $o\isa c[m\fd r]$ is a reference to $o$ and. additionally,
it specifies $o$'s membership in class $c$ and the value of the attribute $m$.

Consequently, all F-logic expressions of the form (0--8) are called
\emph{references}. F-logic references have a dual reading: Given an
\fl\ database \db I (see below), a reference has
\begin{itemize}
\item an \emph{object value}, which yields the name(s) of the objects
  reachable in \db I by the corresponding reference, and 
\item a \emph{truth value} like any other literal or molecule of the
  language; in particular, a reference $r$ evaluates to \emph{false} if
  there exists no object that is referenced by $r$ in \db I.
\end{itemize}
Thus, a path expression can be viewed as a logical formula
(\emph{the deductive perspective}), or as a name for a number of objects
(\emph{the object-oriented perspective}).

Consider the following path expression and the equivalent (with respect to
the truth value) flattening:

\begin{displaymath}
a..b[c\mvd\{d.e\}] \quad\ \Leftrightarrow \quad\  a[b\mvd\{X_{ab}\}]
\land d[e\fd X_{de}] \land X_{ab}[c\mvd\{X_{de}\}]. \hspace{4em} (*)
\end{displaymath}


Such flattening is used to determine the truth value of
arbitrarily complex path expressions in the \emph{body} of a rule.
The object values \obj of a path expression are
the names of the referenced objects: e.g., for $(*)$ we have
\begin{displaymath}
\obj(a..b) = \{x_{ab} \mid \db I \models a[b\mvd\{x_{ab}\}]\}
\qquad\textrm{ and }\qquad \obj(d.e) = \{x_{de} \mid \db I \models d[e\fd 
x_{de}]\} ~,
\end{displaymath}
%
where $\db I \models \varphi$ means that $\varphi$ holds in \db I.
Observe that $\obj(d.e)$ contains at most one element because the
\emph{single-valued} method $e$ is applied to the unique oid $d$.  In
general, for an \fl\ database \db I, the object values of ground
expressions are given by the following mapping \obj\ from ground
references to sets of ground references:
%
\begin{displaymath}
  \begin{array}{cll@{\hspace{4em}}c}
    \obj(t) & := & \{t' \mid t'=t \textrm{ and } \db I\models t' \}, 
     \textrm{ for a ground id-term $t$}  & (0) \\   
                                %
    \obj(o[\dots]) & :=& \{o'\in\obj(o) \mid \db I \models o'[\dots]
    \}& (1),...,(4)\\  
                                %
    \obj(o\isa c) & := & \{o'\in\obj(o) \mid \db I \models o'\isa c\} &
    (5) \\ 
                                %
    \obj(c\subcl d) & := & \{c'\in\obj(c) \mid \db I \models c'\subcl
    d\} &  (6)\\ 
                                %
    \obj(o.m) & :=  & \{r'\in\obj(r) \mid \db I \models o[m\fd
    r]\}  &  (7)\\ 
                                %
    \obj(o..m) & := &  \{ r'\in\obj(r) \mid \db I \models
    o[m{\mvd}\{r\}] \} & (8)   
  \end{array}
\end{displaymath}
Observe that if $t$ does not hold (i.e., occur) in \db I, then $\obj(t)$ is
$\emptyset$.  Conversely, a ground reference $r$ is called \emph{active} if
$\obj(r)\neq\emptyset$. A reference, $r$, can be classified as either
single-valued or multi-valued:
\begin{itemize}
\item $r$ is called \emph{multi-valued} if
 \begin{itemize}
  \item it has the form $o..m$, or 
  \item it has one of the forms $\underline{o}[\dots]$,
    $\underline{o}\isa c$, $\underline{c}\subcl d$, or
    $\underline{o}.\underline{m}$, and any of the underlined
    subexpressions is multi-valued;
 \end{itemize}
\item in all other cases, $r$ is \emph{single-valued}.
\end{itemize}



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manual2"
%%% End: 
