\section{Incremental Table Maintenance} \label{incremental_tabling}
%====================================================

XSB allows the user to declare that the system should incrementally
maintain particular tables.  An incrementally maintained table is one
that continually contains the correct answers in the presence of
updates to underlying predicates on which the tabled predicate
depends.  If tables are thought of as database views, then this
subsystem enables what is known in the database community as
incremental view maintenance.

\subsection{Examples}

To demonstrate incremental table maintenance, we consider first the
following simple program that does not use incremental tabling:
\begin{verbatim}
:- table p/2.

p(X,Y) :- q(X,Y),Y =< 5.

:- dynamic q/2.

q(a,1).
q(b,3).
q(c,5).
q(d,7).
\end{verbatim}
and the following queries and results:
\begin{verbatim}
| ?- p(X,Y),writeln([X,Y]),fail.
[c,5]
[b,3]
[a,1]

no
| ?- assert(q(d,4)).

yes
| ?- p(X,Y),writeln([X,Y]),fail.
[c,5]
[b,3]
[a,1]

no
| ?- 
\end{verbatim}
Here we see that the table for {\tt p/2} depends on the contents of
the dynamic predicate {\tt q/2}.  We first evaluate a query, {\tt
p(X,Y)}, which creates a table.  Then we use {\tt assert} to add a
fact to the {\tt q/2} predicate and re-evaluate the query.  We see
that the answers haven't changed, and this is because the table is
already created and the second query just retrieves answers directly
from that existing table.  But in this case we have answers that are
inconsistent with the current definition of {\tt p/2}.  I.e., if the
table didn't exist (e.g. if {\tt p/2} weren't tabled), we would get a
different answer to our {\tt p(X,Y)} query, this time including the
[d,4] answer.  The usual solution to this problem is for the XSB
programmer to explicitly abolish a table whenever changing (with
assert or retract) a predicate on which the table depends.

By declaring that the tables for {\tt p/2} should be incrementally
maintained, and using specific dynamic predicate update predicates,
the system will automatically keep the tables for {\tt p/2} correct.
Consider the program:
\begin{verbatim}
:- incr_table p/2.

p(X,Y) :- q(X,Y),Y =< 5.

:- incr_dynamic q/2.

q(a,1).
q(b,3).
q(c,5).
q(d,7).
\end{verbatim}
in which {\tt p/2} is declared to be incrementally tabled (with {\tt
:- incr\_table}) and {\tt q/2} is declared to be {\tt :-
incr\_dynamic}, meaning that an incremental table depends on it.
Consider the following goals and execution:
\begin{verbatim}
| ?- import incr_assert/1 from increval.

yes
| ?- p(X,Y),writeln([X,Y]),fail.
[c,5]
[b,3]
[a,1]

no
| ?- incr_assert(q(d,4)).

yes
| ?- p(X,Y),writeln([X,Y]),fail.
[d,4]
[c,5]
[b,3]
[a,1]

no
| ?- 
\end{verbatim}
Here again we call {\tt p(X,Y)} and generate a table for it and its
answers.  (We have imported the {\tt incr\_assert} predicate we need
to interact with the incremental table maintenance subsystem.)  Then
we update {\tt q/2} by using the incremental version of assert, {\tt
incr\_assert/1}.  Now when we call {\tt p(X,Y)} again, the table has
been updated and we get the correct answer.

In this case after every {\tt incr\_assert} and/or {\tt
incr\_retract}, the tables are incrementally updated to reflect the
change.  The system keeps track of what tabled goals depend on what
other tabled goals and (incr) dynamic goals, and tries to minimize the
amount of recomputation necessary.  Incrementally tabled predicates
may depend on other tabled predicates.  In this case, those tabled
predicates must also be declared as incremental (or opaque).  The
algorithm used is described in [].

We note that there is a more efficient way to program incremental
updates when there are several changes made to the base predicates at
one time.  In this case the {\tt incr\_assert\_inval} and {\tt
incr\_retract\_inval} operations should be used for each individual
update.  These operations leave the dependent tables unchanged (and
thus inconsistent.)  Then to update the tables for all the changes
made, the user should call {\tt incr\_table\_update}.

\subsection{Predicates for Incremental Table Maintenance}

The following directives support incremental tabling:

\begin{description}
\ournewitem{incr\_table(+PredSpecs)}{standard}\index{\texttt{incr\_table/1}}
is a compiler directive that indicates that each predicate specified
in {\tt PredSpec} is to be tabled and the tables are to be
incrementally maintained.  {\tt PredSpec} is a list of skeletons,
i.e. open terms, or {\tt Pred/Arity} specifications.  The tables will
be variant tables.  If a predicate is already declared as
subsumptively tabled, an error is thrown.

We also note that any tabled predicate that is called by an {\tt
incr\_table}-d predicate must also be {\tt incr\_table}-d (or {\tt
opaque\_table}-d.)  A dynamic predicate that is called by an {\tt
incr\_table}-d predicate may or may not be declared {\tt
incr\_dynamic}.  But if one is not declared {\tt incr\_dynamic}, then
changes to it will not be propagated to incrementally maintained
tables.

\ournewitem{incr\_dynamic(+PredSpecs)}{standard}\index{\texttt{incr\_dynamic/1}}
is a compiler directive that indicates that a predicate is dynamic and
used to define an incrementally tabled predicate and will be updated
using {\tt incr\_assert} and/or {\tt incr\_retract}.  

\ournewitem{incr\_opaque\_table(+PredSpecs)}{standard}\index{\texttt{incr\_opaque\_table/1}}
is a compiler directive that indicates that a predicate is tabled and
used in the definition of some incrementally tabled predicate but it
should not be maintained incrementally.  In this case the system
assumes that the programmer will abolish tables for this predicate in
such a way so that re-calling it will always give semantically correct
answers.  So instead of maintaining information to support incremental
table maintenance, the system re-calls the opaque predicate whenever
its results are required to recompute an answer.  One example of an
appropriate use of opaque is for tabled predicates in a DCG used to
parse some string.  Rather than incrementally maintain all
dependencies on all input strings, the user can declare these
intermediate tables as opaque and abolish them before any call to the
DCG.

\end{description}

The following predicates are used to manipulate incrementally
maintained tables:

\begin{description}
\ournewitem{incr\_assert(+Clause)}{increval}\index{\texttt{incr\_assert/1}}
is a version of {\tt assert/1} for dynamic predicates declared as
incr\_dynamic.  This adds the clause to the database after any other
clauses for the same predicate currently in the database.  It then
updates all incrementally maintained tables that depend on this
predicate.

\ournewitem{incr\_assertz(+Clause)}{increval}\index{\texttt{incr\_assert/1}}
is the same as {\tt incr\_assert/1}.

\ournewitem{incr\_asserta(+Clause)}{increval}\index{\texttt{incr\_assert/1}}
is the same as {\tt incr\_assert/1} except that it adds the clause
before any other clauses for the same predicate currently in the
database.

\ournewitem{incr\_retract(+Clause)}{increval}\index{\texttt{incr\_retract/1}}
is a version of {\tt retract/1} for dynamic predicates declared as
{\tt incr\_dynamic}.  This removes all clauses in the database that
match Clause.  It then updates all incrementally maintained tables
that depend on this predicate.

\ournewitem{incr\_assert\_inval(+Clause)}{increval}\index{\texttt{incr\_assert\_inval/1}}
is similar to {\tt incr\_assert/1} except that it does not update the
incrementally maintained tables.  The tables should be updated by an
explicit call to {\tt incr\_table\_update/0} (or {\tt /1} or {\tt
/2}).  This separation of function allows for more efficient
processing of table maintenance after a batch of operations.

\ournewitem{incr\_assertz\_inval(+Clause)}{increval}\index{\texttt{incr\_assert\_inval/1}}
is similar to {\tt incr\_assertz/1} except that it does not update the
incrementally maintained tables.  The tables should be updated by an
explicit call to {\tt incr\_table\_update/0} (or {\tt /1} or {\tt
/2}).

\ournewitem{incr\_asserta\_inval(+Clause)}{increval}\index{\texttt{incr\_assert\_inval/1}}
is similar to {\tt incr\_asserta/1} except that it does not update the
incrementally maintained tables.  The tables should be updated by an
explicit call to {\tt incr\_table\_update/0} (or {\tt /1} or {\tt /2}).

\ournewitem{incr\_retract\_inval(+Clause)}{increval}\index{\texttt{incr\_retract\_inval/1}}
is similar to {\tt incr\_retract/1} except that it does not update the
incrementally maintained tables.  The tables should be updated by an
explicit call to {\tt incr\_table\_update/0} (or {\tt /1} or {\tt
/2}).

\ournewitem{incr\_table\_update}{increval}\index{\texttt{incr\_table\_update/0}}
is called after base predicates have been changed (by {\tt
incr\_assert\_inval/1} and/or {\tt incr\_retract\_inval/1} or
friends).  This predicate updates all the incrementally maintained
tables whose contents change as a result of those changes to the base
predicates.  This update operation is separated from the operations
that change the base predicates ({\tt incr\_assert\_inval} and {\tt
incr\_retract\_inval}) so that a set of base predicate changes can be
processed all at once, which may be much more efficient that updating
the tables at every base update.

\ournewitem{incr\_table\_update(-GoalList)}{increval}\index{\texttt{incr\_table\_update/1}}
is similar to {\tt incr\_table\_update/0} in that it updates the
incrementally maintained tables after changes to base predicates.  It
returns the list of goals to incrementally maintained tables whose
tables were changed in the update process.

\ournewitem{incr\_table\_update(+SkelList,-GoalList)}{increval}\index{\texttt{incr\_table\_update/2}}
is similar to {\tt incr\_table\_update/1} in that it updates the
incrementally maintained tables after changes to base predicates.  The
first argument is a list of predicate skeletons (open terms) for
incrementally maintained tables.  The predicate returns in {\tt
GoalList} a list of goals whose skeletons appear in {\tt SkelList} and
whose tables were changed in the update process.  So {\tt SkelList}
acts as a filter to restrict the goals that are returned to those of
interest.  If {\tt SkelList} is a variable or the empty list, all
affected goals are returned in {\tt GoalList}.

\ournewitem{incr\_invalidate\_call(+Goal)}{increval}\index{\texttt{incr\_invalidate\_call/1}}
is used to directly invalidate a call to an incrementally maintained
table.  {\tt Goal} is the tabled call to invalidate.  A subsequent
invocation of {\tt incr\_table\_update} will cause that tabled goal to
be recomputed and all incrementally maintained tables depending on
that goal will be updated.  This predicate can be used if a tabled
predicate depends on some external data and not (only) on {\tt
incr\_dynamic} predicates.  If, for example, an incrementally
maintained predicate depends on a relation stored in an external
relational database (perhaps accessed through the ODBC interface),
then this predicate can be used to invalidate the table when the
external relation changes.  The application programmer must know when
the external relation changes and invoke this predicate as necessary.

\ournewitem{incr\_directly\_depends(?DependentGoal,?Goal)}{increval}\index{\texttt{incr\_directly\_depends/2}}
accesses the dependency structures used by the incremental table
maintenance subsystem to provide information about which incremental
table calls depend on which others.  At least one of {\tt
DependentGoal} and {\tt Goal} must be bound.  If {\tt DependentGoal}
is bound, then this predicate will return in {\tt Goal} through
backtracking the goals for all incrementally maintained tables that
tables unifying (?) with {\tt DependentGoal} directly depend on.  If
{\tt Goal} is bound, then it returns the directly dependent tabled
goals in {\tt DependentGoal}.  [check this out...]

\ournewitem{incr\_trans\_depends(?DependentGoal,?Goal)}{increval}\index{\texttt{incr\_trans\_depends/2}}
is similar to {\tt incr\_directly\_depends} except that it returns
goals according to the transitive closure of the ``directly depends''
relation.
\end{description}
