\chapter{XSB - Oracle Interface} \label{oracle_inter}
%====================================================

\begin{center}
{\Large {\bf By Hassan Davulcu and Ernie Johnson }}
\end{center}

\section{Introduction}
%=====================

The XSB\,-\,Oracle interface provides the programmer with two levels of
interaction.  The first, {\it relation level interface},
offers a tuple-at-a-time retrieval of information from the Oracle
tables.  The second, {\it view level interface}, can translate an
entire Prolog clause into a single SQL query to the Oracle, including
joins and aggregate operations.

This interface allows Oracle tables to be accessed from XSB's
environment as though they existed as facts. All database accesses are
done on the fly allowing XSB to sit alongside other concurrent tasks.

Our interface gives an Oracle programmer all the features of Prolog as
a query language including intensional database specification,
recursion, the ability to deal with incomplete knowledge, inference
control through the {\it cut} operation, and the representation of
negative knowledge through negation.


\subsection{Interface features}
%==============================
\begin{itemize} 
\item Concurrent access for multiple XSB systems to Oracle 7.1.3
	running under Solaris
\item Full data access and cursor transparency including support for
	\begin{itemize}
	\item Full data recursion through XSB's tabling mechanism
	\item Runtime type checking
	\item Automatic handling of NULL values for insertion, 
		deletion and querying
	\item Partial recovery for cursor losses due to cuts
	\end{itemize}
\item Full access to Oracle's SQLplus including
	\begin{itemize}
	\item Transaction support
	\item Cursor reuse for cached SQL statements 
		with bind variables (by avoiding re-parsing and re-declaring).
	\item Caching compiler generated SQL statements with bind variables 
		and efficient cursor management for cached statements
	\end{itemize}
\item A powerful Prolog / SQL compiler based on \cite{Drax92}.
\item Full source code availability for ports to other versions of
      Oracle or other platforms
\item Independence from database schema by employing {\it relation level}
\item Performance as SQL by employing {\it view level} 
\item No mode specification is required for optimized view compilation
\end{itemize}

\section{Installation}
%======================

General information on XSB installation procedures can be found in
Chapter~2 of Volume~1 of the \emph{XSB Programmer's Manual}.  The
following instructions build upon that foundation, and assume that
Oracle component(s) are currently installed.

\paragraph{Unix instructions:}
\begin{enumerate}
\item Set the environment variable {\tt LDFLAGS} to indicate the
Oracle libraries needed to build the system.  For instance:
%%
\begin{quote}
	\texttt{LDFLAGS=-lclntsh -lcommon -lcore4 -lnlsrtl3}
\end{quote}
%%
or
%%
\begin{quote}
	\texttt{setenv LDFLAGS "-lclntsh -lcommon -lcore4 -lnlsrtl3"}
\end{quote}
%%
depending on the shell that you are using.  We have found that the
required libraries are frequently dependent upon the version of
Oracle, the version of any additional Oracle component~---~such as
SQL*Net~---~and possibly the operating system.  Hence, they will
likely differ from those shown.  We further warn that the \emph{order}
in which the libraries are listed may be important.  We have also
witnessed that certain libraries may require multiple listings.
Please refer to your Oracle documentation.

\item Change directory to \texttt{\$XSB\_DIR/build} and run the XSB
configuration script, {\tt configure}, with the options
\begin{quote}
	{\tt --with-oracle}\quad
	{\tt --with-static-libraries=\emph{OracleLibPath}}
\end{quote}
where \texttt{\emph{OracleLibPath}} is the directory that contains
the Oracle client libraries.

With some configurations, compilation of Pro*C files requires that you
use a C compiler other than the default one.  If such is the case,
then additionally pass the option
\begin{quote}
	{\tt --with-cc=\emph{compiler}}
\end{quote}
to \texttt{configure}, indicating that compiler
\emph{\texttt{compiler}} should instead be used.

As a final measure, if Oracle include files cannot be located
automatically, then employ the option
\begin{quote}
	{\tt --with-includes=\emph{OracleHeaderPath}}
\end{quote}
where \texttt{\emph{OracleHeaderPath}} is the directory that contains
the Oracle client header files.

Note the final message that {\tt configure} produces before
completing.  It indicates how you should invoke \texttt{makexsb} to
build an Oracle-enabled version of XSB (step~4).

\item Change directory to \texttt{\$XSB\_DIR/emu} and create the file
\texttt{orastuff.c} from \texttt{orastuff.pc} by invoking the Pro*C
preprocessor:
\begin{quote}
	{\tt make -f proc.mk orastuff.c}
\end{quote}
A sample makefile, \texttt{proc.mk}, is provided, but you should use
the one included with your version of Oracle.  This used to be found
with sample SQL*Plus and Pro*C files in the Oracle installation
hierarchy.

\textbf{Note: this file may provide insight into the library paths and
linking options required by your Oracle installation.}

\item Run {\tt makexsb} as directed by the message displayed at the
end of the run of the \texttt{configure} script (from step~2).

If errors ensue, then review the cause(s) and consider employing the
additional options mentioned in step~2.  If one or more are
appropriate, then apply them, repeating this process from that step.

When {\tt makexsb} completes, it will provide you with the name of the
XSB executable.  Unless you have used the configuration option
\begin{quote}
	{\tt --config-tag=\emph{tag}}
\end{quote}
this will likely be {\tt ../bin/xsb-ora}.
\end{enumerate}

\paragraph{Windows instructions:}
To build XSB with Oracle support, type the following in the {\tt emu}
directory: 
%%
\begin{quote}
 {\tt NMAKE /f "MS\_VC\_Mfile.mak" CFG="release" ORACLE="yes" SITE\_LIBS="libraries"  }
\end{quote}
%%
The {\tt SITE\_LIBS} parameter should include the list of necessary Oracle
support libraries (per Oracle instructions). When the compiler is done, the 
XSB executable is found in its usual place:
%%
\[
 \tt
 \$XSB\_DIR\backslash config\backslash \mbox{\tt x86-pc-windows}\backslash bin
 \backslash xsb.exe
\]
%%


\section{Using the Interface} \label{oracle:use}

Begin by  starting XSB and loading the interface:
\begin{quote}
	\texttt{| ?- [ora\_call].}
\end{quote}

\subsection{Connecting to and disconnecting from Oracle}

Assuming the Oracle server is running, you have an account, and the
environment variables \texttt{ORACLE\_SID} and \texttt{ORACLE\_HOME}
are set, you can login to Oracle by invoking {\tt db\_open/1} as:
\begin{quote}
{\tt  | ?- db\_open(oracle(Name, Password)).}
\end{quote}
If the login is successful, there will be a response of {\tt yes}.

To reach a remote server you can use:
\begin{quote}
	{\tt  | ?- db\_open(oracle('Name@\emph{dblink}', Password)).}
\end{quote}
where {\tt \emph{dblink}} contains the machine name, and optionally
the protocol and server instance name.  For example
\begin{quote}
 {\tt | ?- db\_open(oracle('SCOTT@T:compserv1gw:INST', 'TIGER')).}
\end{quote}
indicates to the runtime system that we want to contact an Oracle
server instance on the host {\tt compserv1gw}, whose {\tt ORACLE\_SID}
is {\tt INST}, using the TCP/IP protocol.  Further, we want to access
that database as the user \texttt{SCOTT} with password \texttt{TIGER}.

To disconnect from the current session use:
\begin{quote}
{\tt  | ?- db\_close.}
\end{quote}


\subsection{Accessing an Oracle Table: Relation Level Interface}
\label{sec:oracle:rellevel}

Assuming you have access permission for the table you wish to import,
you can use {\tt db\_import/2} as:
\begin{quote}
{\tt | ?- db\_import('TABLENAME'('FIELD1', 'FIELD2', .., 'FIELDn'), 'Pname').}
\end{quote}
where {\tt 'TABLENAME'} is the name of the table you wish to access
and {\tt 'Pname'} is the name of the predicate you wish to use to
access the table from XSB. {\tt 'FIELD1'} through {\tt 'FIELDn'} are
the exact attribute names as defined in the database catalog.  The
chosen attributes define the view and the order of arguments for the
database predicate {\tt 'Pname'}.  For example, to create a link to
the {\tt DEPT} table through the {\tt 'dept'} predicate:
\begin{verbatim}
| ?- db_import('DEPT'('DEPTNO','DNAME','LOC'),dept).

yes
| ?- dept(Deptno, Dname, Loc).

Deptno = 10
Dname = ACCOUNTING
Loc = NEW YORK 
\end{verbatim}

Backtracking can then be used to retrieve the next row of the table {\tt DEPT}.

Records with particular field values may be selected in the same way
as in Prolog.  (In particular, no mode specification for database predicates is
required). For example:
\begin{quote}

{\tt | ?- dept(A, 'ACCOUNTING', C).}
\end{quote}
generates the query:
\begin{verbatim}

SELECT DEPTNO, LOC
FROM DEPT rel1
WHERE rel1.DNAME = :BIND1;
\end{verbatim}
and 
\begin{quote}

{\tt | ?- dept('NULL'(\_), 'ACCOUNTING', C).}
\end{quote}
generates: (See section \ref{NULL-values})
\begin{verbatim}
SELECT NULL , rel1.DNAME , rel1.LOC
FROM DEPT rel1
WHERE rel1.DEPTNO IS NULL AND rel1.DNAME = :BIND1;
\end{verbatim}
During the execution of this query the {\tt :BIND1} variable will be bound
to {\tt 'ACCOUNTING'}.\newline
If a field includes a quote $(')$ then this should be represented by
using two quotes.

Note that the relation level interface can be used to define and
access simple project views of single tables.  For example:
\begin{quote}

{\tt | ?- db\_import('DEPT'('DEPTNO','DNAME'),deptview).}
\end{quote}
defines {\tt deptview/2}.

The predicate {\tt db\_import/2} $($and other Oracle interface
predicates$)$ automatically asserts data dictionary information.  You
can use the Prolog predicate {\tt listing/2} to see the asserted data
dictionary information at any time.  

% You do not need to assert any dictionary information at any time.

Note: as a courtesy to Quintus Prolog users we have provided
compatibility support for some PRODBI predicates which access tables
at a relational level.

\begin{verbatim}
i) | ?- db_attach(Pname, table(Tablename)).
\end{verbatim}

eg. execute 
\begin{quote}
{\tt | ?- db\_attach(dept, table('DEPT')).} 
\end{quote}
then execute 
\begin{quote}	
{\tt | ?- dept(Depno, Dname, Loc).}
\end{quote}
to retrieve the rows.
\begin{verbatim}

ii) | ?- db_record('DEPT', R).

    R = [20,RESEARCH,DALLAS];

    R = ...

\end{verbatim}
    You can use {\tt db\_record/2} to treat the whole database row as a single list structure.



\subsection{View Level Interface}


% how about built-ins.

   The view level interface can be used for the definition of rules
whose bodies includes only imported database predicates (by using the
relation level interface) described above and aggregate predicates
(defined below).  In this case, the rule is translated into a complex
database query, which is then executed taking advantage of the query
processing ability of the database system.

One can use the view level interface through the predicate {\tt
db\_query/2}:  
\begin{quote}
{\tt | ?- db\_query('Rulename'(Arg1, ... , Argn), DatabaseGoal).}
\end{quote}
All arguments are standard Prolog terms.  {\tt Arg\_1} through {\tt Arg\_n}
defines the attributes to be retrieved from the database, while
{\tt DatabaseGoal} defines the selection restrictions and join conditions.

The compiler is a simple extension of \cite{Drax92} which generates SQL
queries with bind variables and handles NULL values as described below
(see section~\ref{NULL-values}).  It allows negation, the expression of
arithmetic functions, and higher-order constructs such as grouping,
sorting, and aggregate functions.

Database goals are translated according to the following rules
from \cite{Drax92}:
\begin{itemize}
\item Disjunctive goals translate to distinct SQL queries
	connected through the UNION operator.
\item Goal conjunctions translate to joins.
\item Negated goals translate to negated EXISTS subqueries.
\item Variables with single occurrences in the body are not
	  translated.
\item Free variables translate to grouping attributes.
\item Shared variables in goals translate to equi-join conditions.
\item Constants translate to equality comparisons of an attribute and
	  the constant value.
\item Nulls are translated to {\tt IS NULL} conditions.
\end{itemize}
For more examples and implementation details see the demo in 
{\tt \$XSB\_DIR/examples/xsb\_ora\_demo.P}, and \cite{Drax92}.
 
In the following, we show the definition of a simple join view between the 
two database predicates {\it emp} and {\it dept}.

Assuming the declarations:
\begin{verbatim}

| ?- db_import('EMP'('ENAME','JOB','SAL','COMM','DEPTNO'),emp).

| ?- db_import('DEPT'('DEPTNO','DNAME','LOC'),dept).

use:
	
| ?- db_query(rule1(Ename,Dept,Loc),
	          (emp(Ename,_,_,_,Dept),dept(Dept,Dname,Loc))).
yes

| ?- rule1(Ename,Dept,Loc).
\end{verbatim}

generates the SQL statement:
\begin{verbatim}

SELECT rel1.ENAME , rel1.DEPTNO , rel2.LOC
FROM emp rel1 , DEPT rel2
WHERE rel2.DEPTNO = rel1.DEPTNO;

Ename = CLARK
Dept = 10
Loc = NEW YORK
\end{verbatim}

Backtracking can then be used to retrieve the next row of the view.
\begin{quote}

{\tt | ?- rule1('CLARK',Dept,'NULL'(\_)).}
\end{quote}

generates the SQL statement:
\begin{verbatim}

SELECT rel1.ENAME , rel1.DEPTNO , NULL
FROM emp rel1 , DEPT rel2
WHERE rel1.ENAME = :BIND1 AND rel2.DEPTNO = rel1.DEPTNO AND rel2.LOC IS NULL;
\end{verbatim}

The view interface also supports aggregate functions predicates sum, avg,
count, min and max.  For example
\begin{verbatim}

| ?- db_query(a(X),(X is avg(Sal,A1 ^ A2 ^ A4 ^ A5 ^ emp(A1,A2,Sal,A4,A5)))).


yes.
| ?- a(X).

generates the query :

SELECT AVG(rel1.SAL)
FROM emp rel1;

X = 2023.2

yes
\end{verbatim}


A more complicated example:
\begin{verbatim}

| ?- db_query(harder(A,B,D,E,S), 
                           (emp(A,B,S,E,D),
                            not dept(D,P,C), 
                            not (A = 'CAROL'),
                            S > avg(Sal,A1 ^ A2 ^ A4 ^ A5 ^ A6 ^ A7 ^(
                                    emp(A1,A2,Sal,A4,A5),
                                    dept(A5,A7,A6),
                                    not (A1 = A2))))).


| ?- harder(A,B,D,E,S).
\end{verbatim}

generates the SQL query:
\begin{verbatim}

SELECT rel1.ENAME , rel1.JOB , rel1.DEPTNO , rel1.COMM , rel1.SAL
FROM emp rel1
WHERE NOT EXISTS
       (SELECT * 
        FROM DEPT rel2 
        WHERE rel2.DEPTNO = rel1.DEPTNO) 
   AND rel1.ENAME <> 'CAROL' 
   AND rel1.SAL > 
	(SELECT AVG(rel3.SAL) 
         FROM emp rel3 , DEPT rel4
	 WHERE rel4.DEPTNO = rel3.DEPTNO 
            AND rel3.ENAME <> rel3.JOB);


A = SCOTT
B = ANALYST
D = 50
E = NULL(null1)
S = 2300
\end{verbatim}

All database rules defined by db\_query can be queried with any mode:
For example:
\begin{quote}

{\tt | ?- harder(A,'ANALYST',D,'NULL'(\_),S).}
\end{quote}

generates the query:
\begin{verbatim}

SELECT rel1.ENAME , rel1.JOB , rel1.DEPTNO , NULL , rel1.SAL
FROM emp rel1
WHERE rel1.JOB = :BIND1 AND rel1.COMM IS NULL AND NOT EXISTS
(SELECT *
FROM DEPT rel2
WHERE rel2.DEPTNO = rel1.DEPTNO
) AND rel1.ENAME <> 'CAROL' AND rel1.SAL > 
(SELECT AVG(rel3.SAL)
FROM emp rel3 , DEPT rel4
WHERE rel4.DEPTNO = rel3.DEPTNO AND rel3.ENAME <> rel3.JOB
);


A = SCOTT
D = 50
S = 2300;

no
\end{verbatim}

Notice that at each call to a database relation or rule, the
communication takes place through bind variables.  The corresponding
restrictive SQL query is generated, and if this is the first call with
that adornment, it is cached.  A second call with same adornment would
try to use the same database cursor if still available, without
parsing the respective SQL statement.  Otherwise, it would find an
unused cursor and retrieve the results.  In this way efficient access
methods for relations and database rules can be maintained throughout
the session.

\subsection{Connecting to an SQL query}


It is also possible to connect to any SQL query using the {\tt
db\_sql\_select/2} predicate which takes an SQL string as its input and
returns a list of field values.  For example:
\begin{verbatim}

| ?- db_sql_select('SELECT * FROM EMP',L).

L = [7369,SMITH,CLERK,7902,17-DEC-80,800,NULL,20];

L = etc ...
\end{verbatim}

And you can use db\_sql/1 for any other non-query SQL statement request.  For 
example:
\begin{verbatim}

| ?- db_sql('create table test ( test1 number, test2 date)').

yes
\end{verbatim}

\subsection{Insertions and deletions of rows}

Inserts are communicated to the database {\it array at a time}.
To flush the buffered inserts one has to invoke {\tt flush/0}
at the end of his inserts. 

For setting the size of the {\it input array} See section \ref{Array-sizes}.

Assuming you have imported the related base table using {\tt db\_import/2}, you can insert to
that table by using {\tt db\_insert/2} predicate.  The first argument is
the declared database predicate for insertions and the second argument
is the imported database relation.  The second argument can be declared with
with some of its arguments bound to constants.  For example assuming {\tt empall} is imported through {\tt db\_import}:
\begin{verbatim}

|?- db_import('EMP'('EMPNO','ENAME','JOB','MGR','HIREDATE','SAL','COMM',
	'DEPTNO'), empall).
yes 
| ?- db_insert(emp_ins(A1,A2,A3,A4,A5,A6,A7),(empall(A1,A2,A3,A4,A5,A6,A7,10))).

yes
| ?- emp_ins(9001,'NULL'(35),'qqq',9999,'14-DEC-88',8888,'NULL'(_)).

yes
\end{verbatim}
Inserts the row: 9001,NULL,'qqq',9999,'14-DEC-88',8888,NULL,10
Note that any call to {\tt emp\_ins/7} should have all its arguments bound.

See section \ref{NULL-values} for information about NULL values.

Deletion of rows from database tables is supported by the {\tt
db\_delete/2} predicate.  The first argument is the declared delete
predicate and the second argument is the imported database relation
with the condition for requested deletes, if any.  The condition is
limited to simple comparisons.  For example assuming
{\tt dept/3} is imported as above:
\begin{verbatim}

| ?- db_delete(dept_del(A), (dept(A,'ACCOUNTING',B), A > 10)). 

yes
\end{verbatim}

After this declaration you can use:
\begin{verbatim}

| ?- dept_del(10).
\end{verbatim}

to generate the SQL statement:
\begin{verbatim}

DELETE DEPT rel1 
WHERE rel1.DEPTNO = :BIND1 
      AND rel1.DNAME = 'ACCOUNTING'
      AND rel1.DEPTNO > 10;
\end{verbatim}


Note that you have to commit your inserts or deletes to tables to make
them permanent.  (See section \ref{Transaction-management}).

\subsection{Input and Output arrays}\label{Array-sizes}

To enable efficient {\it array at a time} communication between the XSB client
and the database server we employ {\it input} and  {\it output} buffer areas.

The {\it input} buffer size specifies the size of the array size to be used
during {\it insertions}. The {\it output} buffer size specifies the size of the
array size to be used during {\it queries}. The default sizes of these arrays
are set to 200. The sizes of these arrays can be queried by {\tt stat\_flag/2}
and they can be modified by {\tt stat\_set\_flag/2}. The flag number assigned
for input array length is 58 and  the flag number assigned for output array 
length is 60.

\subsection{Handling NULL values}\label{NULL-values}


The interface treats NULL's by introducing a single valued function
{\tt 'NULL'/1} whose single value is a unique $($Skolem$)$ constant.
For example a NULL value may be represented by 
\begin{verbatim}
	'NULL'(null123245) 
\end{verbatim} 
Under this representation, two distinct NULL values will not unify.
On the other hand, the search condition {\tt IS NULL Field} can be
represented in XSB as {\tt Field = 'NULL'(\_)}

Using this representation of NULL's the following protocol for queries
and updates is established.

\subsubsection{Queries}

\begin{center}

{\tt | ?- dept('NULL'(\_),\_,\_).}
\end{center}

Generates the query: 


\begin{verbatim}


SELECT NULL , rel1.DNAME , rel1.LOC
			  FROM DEPT rel1
			  WHERE rel1.DEPTNO IS NULL;
\end{verbatim}

Hence, {\tt 'NULL'(\_)} can be used to retrieve rows with NULL values 
at any field.

{\tt 'NULL'/1} fails the predicate whenever it is
used with a bound argument.
\begin{center}

{\tt | ?- dept('NULL'(null2745),\_,\_). $\rightarrow$ fails always.}
\end{center}


\subsubsection{Query Results}
When returning NULL's as field values, the interface returns {\tt NULL/1} 
function with a unique integer argument serving as a skolem constant.

Notice that the above guarantees the expected semantics for the join 
statements.  In the following example, even if {\tt Deptno} is NULL for some rows in {\tt emp} or {\tt dept} tables, the query still evaluates the join successfully.
\begin{center}

{\tt | ?- emp(Ename,\_,\_,\_,Deptno),dept(Deptno,Dname,Loc)..}
\end{center}

\subsubsection{Inserts}

To insert rows with NULL values you can use {\tt Field = 'NULL'(\_)} or
{\tt Field = 'NULL'(null2346)}.  For example:

\begin{center}

{\tt | ?- emp\_ins('NULL'(\_), ...).  $\rightarrow$ inserts a NULL value for ENAME}
\end{center}
\begin{center}

{\tt | ?- emp\_ins('NULL'('bound'), ...) $\rightarrow$ inserts a NULL value for ENAME.}

\end{center}


\subsubsection{Deletes}


To delete rows with NULL values at any particular {\tt FIELD} use {\tt Field = 'NULL'(\_)}, {\tt 'NULL'/1} with a free argument.  When {\tt 'NULL'/1} 's argument
is bound it fails the delete predicate always.  For example:

\begin{center}
{\tt | ?- emp\_del('NULL'(\_), ..).  $\rightarrow$ adds ENAME IS NULL to the generated SQL statement}
\end{center}
\begin{center}

{\tt | ?- emp\_del('NULL'('bound'), ...).  $\rightarrow$ fails always}

\end{center}

The reason for the above protocol is to preserve the semantics of deletes, 
when some free arguments of a delete predicate get bound by some preceding
predicates.  For example in the following clause, the semantics is preserved 
even if the {\tt Deptno} field is NULL for some rows.


\begin{center}

{\tt | ?- emp(\_,\_,\_,\_,Deptno), dept\_del(Deptno).}
\end{center}




\subsection{Data dictionary}


The following utility predicates access the data dictionary.  Users of
Quintus Prolog may note that these predicates are all PRODBI
compatible.  The following predicates print out the indicated information:
\begin{description}

\item[db\_show\_schema(accessible)]
	 Shows all accessible table names for the user.  This list can be long!

\item[db\_show\_schema(user)]
	Shows just those tables that belongs to you.

\item[db\_show\_schema(tuples('TABLE'))]
	Shows the contents of the base table named {\tt 'TABLE'}.

\item[db\_show\_schema(arity('TABLE'))]
	The number of fields in the table {\tt 'TABLE'}.

\item[db\_show\_schema(columns('TABLE'))]
	The field names of a table.
\end{description}

For retrieving above information use:
\begin{itemize}

\item db\_get\_schema(accessible,List)
\item db\_get\_schema(user,List)
\item db\_get\_schema(tuples('TABLE'),List)
\item db\_get\_schema(arity('TABLE'),List)
\item db\_get\_schema(columns('TABLE'),List)
\end{itemize}

The results of above are returned in List as a list.


\subsection{Other database operations}

\begin{description}

\item[db\_create\_table('TABLE\_NAME','FIELDS')]
	{\tt FIELDS} is the field specification as in SQL.
\begin{verbatim}

eg. db_create_table('DEPT', 'DEPTNO NUMBER(2),
                             DNAME VARCHAR2(14),
                             LOC VARCHAR2(13)').
\end{verbatim}


\item[db\_create\_index('TABLE\_NAME','INDEX\_NAME', index(\_,Fields))]
	{\tt Fields} is the list of columns for which an index
	is requested.  For example:
\begin{center}

{\tt db\_create\_index('EMP', 'EMP\_KEY', index(\_,'DEPTNO, EMPNO')).}
\end{center}

\item[db\_delete\_table('TABLE\_NAME')] To delete a table named {\tt 'TABLE\_NAME'}

\item[db\_delete\_view('VIEW\_NAME')] To delete a view named {\tt 'VIEW\_NAME'}

\item[db\_delete\_index('INDEX\_NAME')] To delete an index named {\tt 'INDEX\_NAME'}
\end{description}

These following predicates are the supported PRODBI syntax for deleting and inserting rows:
\begin{description}

\item[\mbox{db\_add\_record('DEPT',[30,'SALES','CHICAGO'])}]
	 arguments are a list composed of field values and the table name to
	 insert the row.

\item[\mbox{delete\_record('DEPT', [40,\_,\_])}]
	 to delete rows from {\tt 'DEPT'} matching the list of values mentioned in
	 second argument. 
\end{description}

For other SQL statements use {\tt db\_sql/1} with the SQL statement as the 
     first argument.  For example:
\begin{center}
{\tt db\_sql('grant connect to fred identified by bloggs')).}
\end{center}


\subsection{Interface Flags}
	
   If you wish to see the SQL query generated by the interface use the 
predicate {\tt db\_flag/3}.  The first parameter indicates the function you wish to
change.  The second argument is the old value, and the third argument specifies
the new value.  For example:
\begin{verbatim}

| ?- db_flag(show_query, Old, on).

     Old = off
\end{verbatim}

SQL statements will now be displayed for all your queries $($the default$)$.
To turn it off use {\tt db\_flag(show\_query,on, off)}.


To enable you to control the error behavior of either the interface or
Oracle database use {\tt db\_flag/3} with fail\_on\_error as first argument.
For example:

\begin{description}

\item[\tt | ?- db\_flag(fail\_on\_error, on, off)]

     Gives all the error control to you,
     (default), hence all requests to Oracle returns true.  You have to check
     each action of yours and take the responsibility for your actions.
     (See \ref{SQLCA})

\item[\tt | ?- db\_flag(fail\_on\_error, off, on)]

	Interface fails whenever something goes wrong.
\end{description}


\subsection{Transaction management}\label{Transaction-management}


Normally any changes to the database will not be committed until the user 
disconnects from the database.  In order to provide the user with some control
over this process, db\_transaction/1 is provided. 
\begin{description}
\item[db\_transaction(commit)]
	Commits all transactions up to this point.
\item[db\_transaction(rollback)]
	Rolls back all transactions since the last commit.
\end{description}
Other services provided by Oracle such that {\tt SET TRANSACTION} can be
effected by using {\tt db\_sql/1}.

Note that depending on Oracle's MODE of operation some or all data manipulation 
statements may execute a commit statement implicitly.


\subsection{SQLCA interface} \label{SQLCA}
%-----------------------------------------
You can use {\tt db\_SQLCA/2} predicate to access the {\tt SQLCA} for error 
reporting or other services.
\begin{description}

\item[db\_SQLCA(Comm, Res)]
	Where Comm is any one of the below and Res is the result
		    from Oracle.
\end{description}

\begin{itemize}
\item SQLCODE: The most recent error code
\item SQLERRML: Length of the most recent error msg
\item SQLERRMC: The error msg 
\end{itemize}
\begin{verbatim}
eg. | ?- db_SQLCA('SQLERRD'(2), Rows).
\end{verbatim}
returns in Rows number of rows processed by the most recent statement.

For SQLCAID, SQLCABC,SQLERRP, 'SQLERRD'(0) to 'SQLERRD'(5), 'SQLWARN'(0), to 
'SQLWARN'(5),'SQLEXT' see {\tt ORACLE's C PRECOMPILER} user's manual.


\subsection{Datalog}
%-------------------
You can write recursive Datalog queries with exactly the same
semantics as in XSB using imported database predicates or database
rules.  For example assuming {\tt db\_parent/2} is an imported database
predicate, the following recursive query computes its transitive closure.

\begin{verbatim}

:- table(ancestor/2).
ancestor(X,Y) :- db_parent(X,Y).
ancestor(X,Z) :- ancestor(X,Y), db_parent(Y,Z).
\end{verbatim}


\subsection{Guidelines for application developers} \label{sec:Guide}
%---------------------------------------------------------------
\begin{enumerate}
\item  Try to group your database predicates and use the view level interface
   to generate efficient SQL queries.
\item  Avoid cuts over cursors since they leave cursors open and can cause
   a leak of cursors.
\item Whenever you send a query get all the results sent by the Oracle by
   backtracking to avoid cursor leaks.  This interface automatically closes a 
   cursor only after you retrieve the last row from the active set.
\item Try to use tabled database predicates for cashing database tables.
\end{enumerate}



\section{Demo}
%-------------
A file demonstrating most of the examples introduced here is included
with this installation in the {\tt examples} directory.  Load the
package and call the goal go/2 to start the demo which is self
documenting.  Do not forget to load {\tt ora\_call.P} first.

\begin{verbatim}
| ?- [ora_call].
| ?- [ora_demo].
[ora_demo loaded]

yes
| ?- go(user, passwd).
\end{verbatim}
where user is your account name, and passwd is your passwd.


\section{Limitations} \label{oracle:limitations}
%----------------------
The default limit on open cursors per session in most Oracle
installations is 50, which is also the default limit for our
interface.  There is also a limit imposed by the XSB interface of 100
cursors, which can be changed upon request
\footnote{e-mail xsb-contact@cs.sunysb.edu}.
If your Oracle installation allows more than 50 cursors but
less then 100 then change the line 
\begin{verbatim} 
#define MAX_OPEN_CURSORS 20 
\end{verbatim} 
in XSB\_DIR/emu/orastuff.pc to your new value, and uncomment enough
many cases to match the above number of cursors plus one, in the switch
statements. Currently this number is 21. Then re-build the system. In 
XSB\_DIR/emu/orastuff.pc we provide code for up to 100 cursors. The last
80 of these cursors are currently commented out.


\section{Error msgs}

\begin{description}
\item[ERR - DB: Connection failed] For some reason you can not connect
	to Oracle.
	\begin{itemize}
	\item	Diagnosis: Try to see if you can run sqlplus.
		If not ask your Oracle admin about it.
	\end{itemize}

\item[ERR - DB: Parse error] The SQL statement generated by the
	Interface or the first argument to {\tt db\_sql/1} or 
	{\tt db\_sql\_select/2} can not be parsed by the Oracle.
	The character number is the location of the error. 
	\begin{itemize}
	\item	Diagnosis: Check your SQL statement.  If our interface
		generated the erroneous statement please contact us at
		{\tt xsb-contact@cs.sunysb.edu}.
	\end{itemize}

\item[ERR - DB: No more cursors left] Interface run out of non-active
cursors either because of a leak (See \ref{sec:Guide}) or you have more then MAX\_OPEN\_CURSORS concurrently open
searches.
\begin{itemize}

\item Diagnosis: System fails always with this error.  db\_transaction(rollback) or
	   db\_transaction(commit) should resolve this by freeing all cursors.
	   Please contact us for more help since this error is fatal for your
	   application.
\end{itemize}

\item[ERR - DB: FETCH failed] Normally you should never get this error if the 
interface running properly.
\begin{itemize}

\item Diagnosis: Please contact us at xsb-contact@cs.sunysb.edu
\end{itemize}

\end{description}


\section{Future work}


% probably needs a more careful rewrite.

We plan to write a precompiler to detect base conjunctions (a
sequence of database predicates and arithmetic comparison predicates)
to build larger more restrictive base conjuncts by classical methods of
rule composition, predicate exchange etc. and then employ the view
level interface to generate more efficient queries and programs.
	Also we want to explore the use of tabling for caching of data
and queries for optimization.



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "manual2"
%%% End: 
